Bridging the Gap Between Visual and Analytical Machine Learning Testing

### 2023-10-25 report/icst24.tex
add items to implications

\begin{description}
  \item[RQ1.] \textbf{How frequently are analytical tests formulated from visualisations created to test ML systems?}

@@

\item[RQ2.] \textbf{What patterns are frequently observed in VA pairs used to test ML systems?}

@@

\item[RQ3.] \textbf{Do the assertions capture all the information obtained from the corresponding visualisation?}

@@

The replication package for this study is made public on Figshare\footnote{The replication package can be found at this URL: https://figshare.com/s/9347228011999cba29f5}.

@@

We manually analysed the source code of the visualisation and assertion code cells. And excluded VA pairs that were not related to each other. We encountered several instances where the tracing was not straightforward due to lack of sufficient context from surrounding markdown cells or poor coding practices---such as use of obscure variable names or deeply nested function calls. In such cases, we used one-shot prompt engineering~\cite{liu2023pretrain} with the ChatGPT 4.0 model to understand the code used to create the visualisation. Listing~\ref{lst:prompt} shows the prompt used for such cases. In a few cases, where the explaination of the visualisation code was not sufficient, we used the ChatGPT model to confirm our annotation of the VA pair. Listing~\ref{lst:prompt-follow-up} shows the prompt used in such cases.

\begin{lstlisting}[caption={Prompt used for ChatGPT 4.0 model to understand visualisation code.}, label={lst:prompt}]
Explain the following Python code:

<code for visualisation>

\end{lstlisting}

\begin{lstlisting}[caption={Prompt used for ChatGPT 4.0 model to confirm if a VA pair are related to each other.}, label={lst:prompt-follow-up}]
Is the following Python assertion related to the visualisation code?

<code for assertion>

\end{lstlisting}

@@

\luis{Amother thought: there are still many visualisations without asserts. We believe in each visualisation there is an expectation that has not been formalised and that it will not be easily validated by other contributors. this is already in the motivation but we can frame it as a visible open problem here as well.}

@@

\luis{some thought that is blended here but maybe we want to make it clear. ML testing is a task that requires practitioners to have a strong background in machine learning, software testing, and mathematics (or statistics). This reinforces the need to adjust the education of these professionals and to design tools that aim mitigate the lack of background in some of these directions.}

@@

Additionally, in the results from RQ2\luis{this is RQ3}, we see several examples of assertions that fail to capture all the information gained from its visual counter-part. We argue that this can lead to a mismatch of expectations in future versions of the ML software: for example, another collaborator of the project might run the code with a new version of the dataset or a different learning technique and assume that if the assertions run successfully, the expected patterns from the visualisation are also being met. After delving into some examples, we also observe that deriving assertions from validations is far from trivial: basic visualisation patterns require complex mathematical techniques that might be out of the expertise of the ML developer. For example, the linear correlation observed with the scatter plot of Figure 6\luis{add ref} is clear to any non-expert but it had to be formalised (cf. Listing 11\luis{add ref}) using the metric coefficient of determination bounded to a sound threshold (0.6). Hence, we ought to bridge the gap between qualitative assessment of ML systems using visualisations and deriving analytical assertions from the visualisations. In other words, there is a need for automated tools that make this translation process easier for AI practitioners. In sum, our results showcase a gap in the current state of the art in software testing: it is failing to address the idiosyncrasies of ML code.

@@

The empirical analysis presented in this study uses Jupyter notebooks mined from Github. To the best of our knowledge,~\cite{pimentel2019large} and~\cite{quaranta2021kgtorrent} are the only two prior studies which provide a public dataset of Jupyter Notebooks. We faced several technical challenges while trying to access the notebooks provided by these authors and thus decided to mine our own data from Github. However, 24\% of the notebooks had to be excluded from the study since they were not related to ML or DS (Section~\ref{sec:identify-related-pairs} and Table~\ref{tab:exclusion-criteria}). We wish to extend our analysis to notebooks collected from Kaggle since Kaggle is predominantly targetted towards the ML and DS community.

In Section~\ref{sec:data-collect}, we intentionally keep the search criteria on Github broad by only checking for the appropriate filetype and the presence of the ``assert'' keyword. Besides being used to define tests in the Python programming language, ``assert'' also appears in several external python libraries that provide a testing interface (for instance nose~\footnote{http://readthedocs.org/docs/nose/}, unittest~\footnote{https://docs.python.org/3/library/unittest.html} and numpy~\footnote{https://numpy.org/doc/stable/reference/routines.testing.html}). Our approach captures a rich variety of assertion statements while preventing the need to craft a custom search query based on an enhaustive list of keywords that appear in several Python testing libraries.

>>>>

\begin{description}
\item[RQ1.] How frequently are analytical tests formulated from visualisations created to test ML systems?

@@

\item[RQ2.] What patterns are frequently observed in VA pairs used to test ML systems?

@@

\item[RQ3.] Do the assertions capture all the information obtained from the corresponding visualisation?

@@

@@

We manually analysed the source code of the visualisation and assertion code cells. And excluded VA pairs that were not related to each other. We encountered several instances where the tracing was not straightforward due to lack of sufficient context from surrounding markdown cells or poor coding practices---such as use of obscure variable names or deeply nested function calls. In such cases, we used one-shot prompt engineering~\cite{liu2023pretrain} with the ChatGPT 4.0 model to determine if the visualisation and assertion were related to each other.

@@

@@

@@

Additionally, in the results from RQ2\luis{this is RQ3}, we see several examples of assertions that fail to capture all the information gained from its visual counter-part. We argue that this can lead to a mismatch of expectations in future versions of the ML software: for example, another collaborator of the project might run the code with a new version of the dataset or a different learning technique and assume that if the assertions run successfully, the expected patterns from the visualisation are also being met. After delving into some examples, we also observe that deriving assertions from validations is far from trivial: basic visualisation patterns require complex mathematical techniques that might be out of the expertise of the ML developer. We ought to bridge the gap between qualitative assessment of ML systems using visualisations and deriving analytical assertions from the visualisations. In other words, there is a need for automated tools that make this translation process easier for AI practitioners. In sum, our results showcase a gap in the current state of the art in software testing: it is failing to address the idiosyncrasies of ML code.

@@

The empirical analysis presented in this study uses Jupyter notebooks mined from Github. To the best of our knowledge, \cite{pimentel2019large} and \cite{quaranta2021kgtorrent} are the only two prior studies which provide a public dataset of Jupyter Notebooks. We faced several technical challenges while trying to access the notebooks provided by these authors and thus decided to mine our own data from Github. We invite the scientific community to extend our catalogue of visual and analytical testing patterns, using the corpus of notebooks provided by the above authors.

In Section~\ref{sec:data-collect}, we intentionally keep the search criteria on Github broad by only checking for the appropriate filetype and the presence of the ``assert'' keyword. Besides being used to define tests in the Python programming language, ``assert'' also appears in several external python libraries that provide a testing interface (for instance nose, unittest and numpy). Our approach provides captures a rich variety of assertion statements while preventing the need to craft a custom search query based on an enhaustive list of keywords that appear in all Python testing libraries.

% TODO don't mention quantity of random sample; then we need to explain/defend this choice
We used an iterative approach to determine the effectiveness of each filter that was used. We applied each filter in the same order as presented in Table~\ref{tab:filters}, starting with F1. We randomly sampled 50 notebooks from the population of notebooks obtained after applying the filter. We then conduct the qualitative assessment presented in Section~\ref{REFME} and observe the number of related visualisation and assertion pairs. In each new iteration, we progressively add the other filters and include the filter if the number of relevant visualisation and assertion pairs increases.

% TODO defend use of one-shot prompt engineering for identifying related visualisation and assertion code cells; how did we mitigate this threat?
